Static Race Detection in OS Kernels by Mining Locking Rules

### 2023-07-31 1.introduction.tex
update

\begin{itemize}
	\item We analyze the challenges of data race detection in OS kernels, and 
	propose three key techniques to address	these challenges: (T1) an {\em 
	alias-aware rule mining method} to automatically deduce locking rules; (T2) 
	a {\em lock-usage analysis} to filter out false positives caused by 
	concurrency; (T3) a {\em pattern-based estimation} to extract harmful data 
	races that can trigger memory or logical bugs such as null-pointer 
	dereferences and data inconsistency.
	\item Based on these three key techniques, we design a practical static 
	analysis approach named \sys, to effectively detect data races and 
	estimate the harmfulness of these data races in OS kernels.
	\item We have evaluated \sys on Linux 6.2, and find xxx data races, 	
	with a false positive rate of xxx. Among these data races, xxx are 
	estimated to be harmful. We have reported these harmful bugs to Linux 
	kernel developers, and xxx of them have been confirmed.
\end{itemize}

>>>>

\begin{itemize}
	\item We analyze the challenges of data race detection in OS kernels, and 
	propose three key techniques to address	these challenges: (T1) an {\em 
	alias-aware rule mining method} to automatically deduce locking rules; (T2) 
	a {\em lock-usage analysis} to filter out false positives caused by 
	code paths that can not execute concurrently; (T3) a {\em pattern-based 
	estimation} to extract harmful data races that can trigger memory or 
	logical bugs such as null-pointer dereferences and data inconsistency.	
	\item Based on these three key techniques, we design a practical static 
	analysis approach named \sys, to effectively detect data races and 
	estimate the harmfulness of these data races in OS kernels.
	\item We have evaluated \sys on Linux 6.2, and find xxx data races, 	
	with a false positive rate of xxx. Among these data races, xxx are 
	estimated to be harmful. We have reported these harmful bugs to Linux 
	kernel developers, and xxx of them have been confirmed.
\end{itemize}

### 2023-07-31 5.evaluation.tex
update

\PP{Unprotected write in the LPFC SCSI driver.} In 
Figure~\ref{fig_case_bugs}(c), the value of the shared variable {\em 
phba->fcf.fcf\_flag} is set to 0 by the function {\em 
lpfc\_unregister\_fcf\_rescan()}, without acquiring the lock {\em 
phba->hbalock}, and this can cause double-fetch in several program sites. For 
example, if {\em phba->fcf.fcf\_flag} is set to 0 by Thread 2 between the two 
accesses to it in the function {\em lpfc\_register\_fcf()} in Thread 1 (Lines 
1873 and 1874), the values of the two accesses can differ, causing 
unpredictable behavior.

>>>>

\PP{Unprotected write in the LPFC SCSI driver.} In 
Figure~\ref{fig_case_bugs}(c), the value of the shared variable {\em 
phba->fcf.fcf\_flag} is set to 0 by the function {\em 
lpfc\_unregister\_fcf\_rescan()}, without acquiring the lock {\em 
phba->hbalock}, and this can cause double-fetch in several program sites. For 
example, if {\em phba->fcf.fcf\_flag} is set to 0 by Thread 2 between the two 
accesses to it in the function {\em lpfc\_register\_fcf()} in Thread 1 (Lines 
1873 and 1874), the values of the two accesses can be different, causing 
unpredictable behaviors.

### 2023-07-31 6.discussion.tex
update

>>>>

\PP{Benefits in other aspects.} \sys uses an alias-aware mining method to 
deduce precise locking rules about whether accesses to a variable should be 
protected by a lock, and which lock are required. Such rules can benefit in 
various aspects.

First, with these rules, developers can improve reliability 
and performance of kernel code. On the one hand, developers can carefully check 
whether necessary lock are used when writing code, and avoid serious bugs 
caused by data races in the early developing stage. On the other hand, 
developers can limit the critical zone protected by an lock/unlock pair to the 
exact accesses that should be protected, and let other code can execute 
concurrently. And these locking rules can also help improve documents of OS 
kernels, which is benefit for long term evolution of kernel codes.

Second, key fields extract \sys are data structure fields that is protected by 
some locks, and this means that these fields is likely to be concurrently 
accessed by different threads. Therefore, key fields can be used in detecting 
various types of concurrency bugs such as use-after-free and 
uninitialized-variable access. Take concurrency use-after-free as an example, 
if a key field is accessed in one thread and freed in another thread, a 
concurrency use-after-free can occur.

Finally, the locking rules deduced by our alias-aware rule mining method can 
enhance other annotation-based analysis such as Clang thread safety 
analysis~\cite{ClangThreadSafety}. Specifically, developers can use 
instrumentation to annotate necessary locks for specific variables in the code 
automatically, according to the locking rules generated by \sys, and run other 
annotation-based analysis to detect data races.

\PP{Comparison to other approaches.} We have try our best to experimentally 
compare \sys to existing works in data race detection including 
RELAY~\cite{Voung:FSE07}, WHOOP~\cite{Deligiannis:ASE15} and 
Goblint~\cite{Vojdani:ASE16}. Unfortunately, they are too old and generate too 
many errors when compiling and running, and thus we compare \sys to them 
methodologically.

\PP{Limitations and future works.} \sys can be improved in some aspects. First, 
our alias-aware mining method only considers accessed variables and protecting 
locks that exist in the same data structure, and can miss some locking rules 
involves special locks such as RCU locks. To overcome this limitation, we plan 
to create a common virtual node for RCU lock in the alias graph, and regard it 
as the ancestor of all other nodes. In this way, all variables have a common 
ancestor with the RCU lock, and can be inferred to be protected by the RCU 
lock, because the accessed variable and the RCU lock are existing in the 
identical data structure by our alias-aware rule mining method. Second, \sys 
does not handle function-pointer calls, and thus it can not build complete call 
graphs, and thus can miss data races reached from function-pointer calls. To 
relieve this limitation, we plan to apply existing function-pointer 
analysis~\cite{Hind:TOPLAS99, Zhang:ISPEC15, Heintze:PLDI01} in \sys, to detect 
more data races in functions that called through function pointers and reduce 
false negatives. Third, \sys performs a path-based analysis and suffers from 
high time overhead, to accelerate its analysis, we plan to employ summer-based 
analysis~\cite{Bai:ATC19, Bai:ATC22} to avoid re-analysis of the same function 
definition. Finally, we plan to apply \sys to detect data races in other OS 
kernels such as FreeBSD~\cite{FreeBSD} and NetBSD~\cite{NetBSD}, and use it to 
detect more types of concurrency bugs including use-after-free and 
uninitialized-variable access.

### 2023-07-31 7.related.tex
update

>>>>

\subsection{Static Analysis of Data Races}
\label{subsec_static}
Many approaches~\cite{Boyapati:OOPSLA02, Anderson:PLDI08, Anderson:PLDI09, 
Zhou:MICRO19, Flanagan:PASTE01, Flanagan:PLDI00, Sadowski:PLATEAU14, 
ClangThreadSafety, Blackshear:OOPSLA18, Choi:PLDI02, Engler:SOSP03, 
Voung:FSE07, Pratikakis:PLDI06, Naik:PLDI06} use static analysis to detect data 
races, without running the checked program. Some of 
them~\cite{Boyapati:OOPSLA02, Anderson:PLDI08, Anderson:PLDI09, Zhou:MICRO19, 
Flanagan:PASTE01, Flanagan:PLDI00, Sadowski:PLATEAU14, ClangThreadSafety} 
perform annotation-based analysis, and request developers to provide locking 
rules about which lock is required when a specific variable is accessed, and 
then detect data races that violate the provided rules. Clang thread safety 
analysis~\cite{ClangThreadSafety} needs developers to annotate code with a {\tt 
GUARDED\_BY} attribute to indicate which lock is required for a specific 
variable, and then works like a type system for multi-threaded 
programs~\cite{Vu:SoICT12, Boyapati:OOPSLA02, Marion:TAMC14} to detect data 
races that violate the provided attribute. Flanagan et 
al.~\cite{Flanagan:PLDI00} design a type system and relay on developers to 
provide additional type annotations to associate a protecting lock with each 
field declaration, and tracks the set of locks held at each program point. And 
then it identifies whether a given program is race free by checking whether 
necessary locks are indeed held at each program site. However, such 
annotation-based approaches are not suitable for race detection in OS kernels, 
because even an expert developer can not provide accurate annotations, because 
the locking rules of kernel code are not well documented, and the logic of 
kernel code tends to be very complex. Other 
approaches~\cite{Blackshear:OOPSLA18, Choi:PLDI02, Engler:SOSP03, Voung:FSE07, 
Pratikakis:PLDI06, Naik:PLDI06} employ lockset-based analysis to detect data 
races automatically. They deduce locking rules based on statistic instead of 
relying on manual annotations. But they does not consider alias 
relationships~\cite{Voung:FSE07, Engler:SOSP03} or just employ imprecise 
flow-insensitive alias analysis~\cite{Choi:PLDI02, 	Pratikakis:PLDI06, 
Naik:PLDI06}. RacerX~\cite{Engler:SOSP03} performs a lockset 
analysis to to detect data races in three modes, including simple checking, 
simple statistical and precise statistical from least to most precise. However, 
even if for the most precise mode, RacerX does not employ any alias analysis, 
and thus introduce many false positives and negatives.

Different from the above static analysis, \sys employs the alias graph to mine 
locking rules automatically. Benefit from the path-based and field-sensitive 
alias relationships generated by alias graph, \sys can deduce locking rules 
effectively. With these locking rules, \sys can detect data races accurately. 
Moreover, \sys exploits a pattern-based estimation to extract harmful data 
races automatically, and can help developers focus on data races that can 
indeed introduce serious issues.

\subsection{Dynamic Analysis of Data Races}
\label{subsec_dynamic}
Runtime analysis does not suffer from the pointer alias problems, and thus many 
approaches~\cite{Lochmann:EuroSys19, Lu:SOSP07, Lu:FSE18, Joshi:ASE08, 	
Liu:NSDI07, Ocallahan:PPoPP03, Serebryany:WBIA09, Marino:PLDI09} detect data 
races through dynamic analysis. LockDoc~\cite{Lochmann:EuroSys19} records 
accesses to variables and lock acquisitions of an instrumented Linux kernel in 
logs, and then infers locking rules from the logs. After that, LockDoc 
automatically detect data races by checking whether a given variable access 
violates the inferred locking rules. O'callahan et al.~\cite{Ocallahan:PPoPP03} 
combine lockset-based detection and happens-before-based detection to detect 
data races, which can get fewer false positives and less overhead than previous 
dynamic analysis. In order to reduce overhead of runtime overhead further, 
LiteRace~\cite{Marino:PLDI09} employs a sampling-based approach based on the 
hypothesis that data races are likely to occur when a thread is executing
an infrequently accessed region in the program. It instruments sampled memory 
accesses, and perform happens-before-based analysis to detect data races, 
according the logs generated by the instrumentation.

However, dynamic analysis suffers from low code coverage, and thus the locking 
rules inferred through logs can be imprecise. Besides, dynamic analyses need to 
run OS kernels with instrumentation and are hard to deploy.

### 2023-07-31 8.conclusion.tex
update

>>>>

In this paper, we develop a novel static approach named \sys to detect data 
races in OS kernels by mining locking rules. It consists of three key 
techniques, including an alias-aware rule mining method to automatically deduce 
locking rules, a lock-usage analysis to filter out false positives caused by 
code paths that can not execute concurrently and a pattern-based estimation to 
extract harmful data races that can trigger memory or logical bugs such as 
null-pointer dereferences and data inconsistency. In the evaluation, \sys finds 
88 real harmful data races, and 32 of them have been confirmed by the 
developers.
